# Trajectory graphs and trend maps {#GraphMap}

Now that we have calculated values for the trends and trajectories, bbsBayes also includes functions to plot and map the values.

```{r GM1, cache=FALSE,echo=TRUE}
# load the trend and indices objects for Scarlet Tanager
load("Scarlet_Tanager_trends_indices.RData")
```

## Graphing the trajectories (annual indices)

The `plot_indices()` function produces a list of ggplot figures that can be combined into a single pdf file:
``` {r GM2_a, eval = FALSE}
# Not run
tp = plot_indices(indices = indices,
                         species = "Scarlet Tanager")
pdf(file = "Scarlet Tanager Trajectories.pdf")
print(tp)
dev.off()
```

Or, we can print plots to individual devices:

``` {r GM2_b}
tp = plot_indices(indices = indices,
                         species = "Scarlet Tanager")
print(tp[[1]])
```

`plot_indices()` also allows the user to add points to show the observed mean counts as well as stacked dots to indicate the number of observations in each year.

``` {r GM3}
tp2 = plot_indices(indices = indices,
                         species = paste("Scarlet Tanager","Full"),
                  add_observed_means = TRUE,
                  add_number_routes = TRUE)

```

And we can compare these indices plots with those generated using only the smooths.

``` {r GM4}
tp3 = plot_indices(indices = indices_smooth,
                         species = paste("Scarlet Tanager","Smooth"),
                  add_observed_means = TRUE,
                  add_number_routes = TRUE)

```

Using the patchwork library, we can show these plots side-by-side:
```{r GM5}
library(patchwork)
print(tp2[[1]]+tp3[[1]])
```

## Adding elements to indices plots

The output of the `plot_indices()` function is a list of ggplot objects, it is relatively easy to add components to the basic plot. 

Here, we'll use the first plot (continental trajectory) from the full population trajectory version above `tp2[[1]]` as the base plot, and we'll add the smooth population trajectory information to the graph in a different colour.

First, extract the smooth population trajectory information in a dataframe.

```{r GM5a}
library(tidyverse)
continental_smooth <- indices_smooth$data_summary %>% 
  filter(Region == "Continental")
  
```
Then we overlap the original plot with a new line and uncertainty ribbon.

```{r GM5b}

overlay_plot <- tp2[[1]] +
  geom_ribbon(data = continental_smooth,inherit.aes = FALSE,
              aes(x = Year,ymax = Index_q_0.25,ymin = Index_q_0.975),
              alpha = 0.2,
              fill = "darkgreen")+
  geom_line(data = continental_smooth, inherit.aes = FALSE,
            aes(x = Year,y = Index),
            colour = "darkgreen")
print(overlay_plot)
```



## Mapping the trends


The trends can be mapped to produce strata maps coloured by species population trends.

``` {r GM6}
mp = generate_map(trends_smooth,
                  select = TRUE,
                  stratify_by = "bbs_cws",
                  species = "Scarlet Tanager Smooth")
print(mp)
```

## Geofacet Trajectories


For stratifications that can be compiled by political regions (i.e., `bbs_cws`, `bbs_usgs`, or `state`), the function `geofacet_plot` will generate a ggplot object that plots the state and province level population trajectories in facets arranged in an approximately geographic arrangement. These plots offer a concise, range-wide summary of a species' population trajectory.

``` {r GM7}
  gf <- geofacet_plot(indices_list = indices_smooth,
                     select = TRUE,
                     stratify_by = "bbs_cws",
                     multiple = TRUE,
                     trends = trends_smooth,
                     slope = F,
                     species = "Scarlet Tanager Smooth")
  
  print(gf)
```






