# Advanced options {#Adv}


## Custom regional summaries

Yes, you can calculate the trend and trajectories for custom combinations of strata, such as a formal trend estimate for populations of Barn Swallow in the South East portion of their range (e.g., BCRs 21,24,25,26,27,35,36, and 37).

``` {r A1, echo = FALSE}
load("data/gamye_Barn_Swallow/jags_mod_full.RData") #saved Barn Swallow model output and data
```


### Define the custom regions as a collection of the existing strata.

First extract a dataframe that defines the original strata used in the analysis.

``` {r A3}
st_comp_regions <- get_composite_regions(strata_type = "bbs_cws")
knitr::kable(head(st_comp_regions))
```

Then add a column to the dataframe that groups the original strata into the desired custom regions.

``` {r A4}
st_comp_regions$SouthEast <- ifelse(st_comp_regions$bcr %in% c(21,24:27,35:37),"SouthEast","Other")
```

### Use defined regions to estimate indices and trends

st_comp_regions can now be used as the dataframe input to the argument alt_region_names in `generate_indices()`, with "SouthEast" as the value for the argument regions. The relevant trends can be calculated using just the `generate_trends()` function.

``` {r A5}
library(patchwork)
custom_indices <- generate_indices(jags_mod = jags_mod,
                                      jags_data = jags_data,
                                      alt_region_names = st_comp_regions,
                                      regions = "SouthEast")
tp = plot_indices(indices = custom_indices,
                         species = "Barn Swallow",
                  add_observed_means = TRUE,
                  add_number_routes = TRUE,
  title_size = 20,
  axis_title_size = 12,
  axis_text_size = 12)
```

``` {r A5plot,fig.height=9}
print(tp[[1]])
print(tp[[2]])

```

Calculate trends

``` {r A5trend, eval=FALSE}
custom_trends <- generate_trends(indices = custom_indices)
knitr::kable(custom_trends[,c(1,3,8,9,14)])
```




## Exporting the JAGS model

You can easily export any of the bbsBayes models to a text file.

``` {r A6, eval=FALSE}
model_to_file(model = "slope",
              filename = "my_slope_model.txt")
```

This is the best way to confirm all of the priors, and model structures.

## Customizing the JAGS model and data

You can also modify the model text (e.g., try a different prior) and run the modified model

``` {r A7, eval=FALSE}
run_model <- function(... ,
                      model_file_path = "my_modified_slope_model.txt",
                      ... )
```


Or, you could add covariates.
For example a GAM smooth to estimate the effect of the day of year on the observations, or an annual weather covariate, or... Then add the relevant covariate data to the jags_data object, and you're off! 

Here's a GitHub repo with an example of a modified `bbsBayes` model. The modified model includes a series of GAM smooths to account for variation in counts over the course of the BBS observation season (mid-May through early July). In addition these seasonal smooths are allowed to vary by decade to account for possible shifts in phenology through time.

https://github.com/AdamCSmithCWS/BBS_seasonal_shift_ALHU

## Comparing Models

Finally, bbsBayes can be used to run cross-validations. For example, the `get_final_values()` function is useful to provide an efficient starting point for a cross-validation runs, without having to wait for another full burn-in period.

Paper that includes an example of how to implement a cross-validation using bbsBayes.

Ornithological Applications: <https://doi.org/10.1093/ornithapp/duaa065> 
Supplement: <https://zenodo.org/badge/latestdoi/228419725>

NOTE: although bbsBayes includes functions to calculate WAIC, recent work has shown that WAIC performs very poorly with the BBS data (<https://doi.org/10.1650/CONDOR-17-1.1>). We recommend a k-fold cross-validation approach, as in the above zenodo archive.